---
title: "Characterization Installation Guide"
author: "Jenna Reps"
date: '`r Sys.Date()`'
header-includes:
    - \usepackage{fancyhdr}
    - \pagestyle{fancy}
    - \fancyhead{}
    - \fancyhead[CO,CE]{Installation Guide}
    - \fancyfoot[CO,CE]{Characterization Package Version `r    utils::packageVersion("Characterization")`}
    - \fancyfoot[LE,RO]{\thepage}
    - \renewcommand{\headrulewidth}{0.4pt}
    - \renewcommand{\footrulewidth}{0.4pt}
output:
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
  html_document:
    number_sections: yes
    toc: yes
---
<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Installation_guide}
-->

# Introduction
This vignette describes how you need to install the Observational Health Data Sciencs and Informatics (OHDSI) [`Characterization`](http://github.com/OHDSI/Characterization) package under Windows, Mac, and Linux.

# Software Prerequisites

## Windows Users
Under Windows the OHDSI Characterization package requires installing:

* R (https://cran.cnr.berkeley.edu/ ) \- (R >= 3.3.0, but latest is recommended)
* Rstudio (https://www.rstudio.com/  )
* Java (http://www.java.com )
* RTools (https://cran.r-project.org/bin/windows/Rtools/) 

## Mac/Linux Users
Under Mac and Linux the OHDSI Characterization package requires installing:

* R (https://cran.cnr.berkeley.edu/ ) \- (R >= 3.3.0, but latest is recommended)
* Rstudio (https://www.rstudio.com/  )
* Java (http://www.java.com )
* Xcode command line tools(run in terminal: xcode-select --install) [MAC USERS ONLY]


# Installing the Package
The preferred way to install the package is by using `remotes`, which will automatically install the latest release and all the latest dependencies.

If you do not want the official release you could install the bleading edge version of the package (latest develop branch).

Note that the latest develop branch could contain bugs, please report them to us if you experience problems.

## Installing Characterization using remotes
To install using `remotes` run:
```{r, echo = TRUE, message = FALSE, warning = FALSE,tidy=FALSE,eval=FALSE}
install.packages("remotes")
remotes::install_github("OHDSI/FeatureExtraction")
remotes::install_github("OHDSI/Characterization")
```

When installing make sure to close any other Rstudio sessions that are using `Characterization` or any dependency.  Keeping Rstudio sessions open can cause locks that prevent the package installing.


# Installation issues
Installation issues need to be posted in our issue tracker:
http://github.com/OHDSI/Characterization/issues

The list below provides solutions for some common issues:

1. If you have an error when trying to install a package in R saying **'Dependancy X not available ...'** then this can sometimes be fixed by running `install.packages('X')` and then once that completes trying to reinstall the package that had the error.

2. I have found that using the github `remotes`` to install packages can be impacted if you have **multiple R sessions** open as one session with a library open can cause the library to be locked and this can prevent an install of a package that depends on that library.


# Acknowledgments

Considerable work has been dedicated to provide the `Characterization` package.

```{r tidy=TRUE,eval=TRUE}
citation("Characterization")
```
